Reproducibility of research during COVID-19: examining the case of population density and the basic reproductive rate from the perspective of spatial analysis

### 2022-10-08 R0-Density-Reanalysis/R0-Density-Reanalysis.tex
Updated with final bibliographical information and created a renv reproducible environment

% Pandoc citation processing
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
% for Pandoc 2.8 to 2.10.1
\newenvironment{cslreferences}%
  {}%
  {\par}
% For Pandoc 2.11+
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1 \everypar{\setlength{\hangindent}{\cslhangindent}}\ignorespaces\fi
  % set entry spacing
  \ifnum #2 > 0
  \setlength{\parskip}{#2\baselineskip}
  \fi
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

% Pandoc header
\usepackage{setspace}\onehalfspacing
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}

\begin{document}
\begin{frontmatter}

  \title{Reproducibility of research during COVID-19: examining the case
of population density and the basic reproductive rate from the
perspective of spatial analysis}
    \author[McMaster University]{Antonio Paez\corref{1}}
   \ead{paezha@mcmaster.ca} 
      \address[McMaster University]{School of Earth, Environment and
Society, 1280 Main St West, Hamilton, Ontario L8S 4K1 Canada}
      \cortext[1]{Corresponding Author}

@@

The emergence of the novel SARS-CoV-2 coronavirus in 2019, and the
global pandemic that followed in its wake, led to an explosive growth of
research around the globe. According to Fraser et al. (2021), over
125,000 COVID-19-related papers were released in the first ten months
from the first confirmed case of the disease. Of these, more than 30,000
were shared in pre-print servers, the use of which also exploded in the
past year (Añazco et al., 2021; Kwon, 2020; Vlasschaert et al., 2020).

Given the ruinous human and economic cost of the pandemic, there has
been a natural tension in the scientific community between the need to
publish research results quickly and the imperative to maintain
consistently high quality standards in scientific reporting; indeed, a
call for maintaining the standards in published research termed the
deluge of COVID-19 publications a ``carnage of substandard research''
(Bramstedt, 2020). Part of the challenge of maintaining quality
standards in published research is that, despite an abundance of
recommendations and guidelines (e.g., Broggini et al., 2017; Brunsdon
and Comber, 2020; Ince et al., 2012; Ioannidis et al., 2014), in
practice reproducibility has remained a lofty and somewhat aspirational
goal (Konkol et al., 2019; Konkol and Kray, 2019). As reported in the
literature, only a woefully small proportion of published research was
actually reproducible before the pandemic (Iqbal et al., 2016; Stodden
et al., 2018), and the situation does not appear to have changed
substantially since (Gustot, 2020; Sumner et al., 2020).

The push for open software and data (Arribas-Bel et al., 2021; e.g.,
Bivand, 2020), along with more strenuous efforts towards open,
reproducible research, is simply a continuation of long-standing
scientific practices of independent verification. Despite the (at times
disproportionate) attention that high profile scandals in science tend
to elicit in the media, science as a collective endeavor is remarkable
for being a self-correcting enterprise, one with built-in mechanisms and
incentives to weed out erroneous ideas. Over the long term, facts tend
to prevail in science. At stake is the shorter-term impacts that
research may have in other spheres of economic and social life. The case
of economists Reinhart and Rogoff comes to mind: by the time the
inaccuracies and errors in their research were uncovered (see Herndon et
al., 2014), their claims about debt and economic growth had already been
seized by policy-makers on both sides of the Atlantic to justify
austerity policies in the aftermath of the Great Recession of
2007-2009\footnote{Nobel Prize in Economics Paul Krugman noted that
  ``Reinhart--Rogoff may have had more immediate influence on public
  debate than any previous paper in the history of economics''
  \url{https://www.nybooks.com/articles/2013/06/06/how-case-austerity-has-crumbled/?pagination=false}}.
As later research has demonstrated, those policies cast a long shadow,
and their sequels continued to be felt for years (Basu et al., 2017).

In the context of COVID-19, a topic that has grabbed the imagination of
numerous thinkers has been the prospect of life in cities after the
pandemic (e.g., Florida et al., 2020); as a result, the implications of
the pandemic for urban planning, design, and management are the topic of
ongoing research (e.g., Sharifi and Khavarian-Garmsir, 2020). The fact
that the worst of the pandemic was initially felt in dense population
centers such as Wuhan, Milan, Madrid, and New York, unleashed a torrent
of research into the associations between density and the spread of the
pandemic. The answers to some important questions hang on the results of
these research efforts. For example, are lower density regions safer
from the pandemic? Are de-densification policies warranted, even if just
in the short term? In the longer term, will the risks of life in high
density regions presage a flight from cities? And, what are the
implications of the pandemic for future urban planning and practice?
Over the past year, numerous papers have sought to throw light on the
underlying issue of density and the pandemic; nonetheless the results,
as will be detailed next, remain mixed. Further, to complicate matters,
precious few of these studies appear to be sufficiently open to support
independent verification.

The objective of this paper is to illustrate the importance of
reproducibility in research in the context of the flood of COVID-19
papers. For this, I focus on a recent study by Sy et al. (2021) that
examined the correlation between the basic reproductive number of
COVID-19, \(R_0\), and population density. The basic reproductive number
is a summary measure of contact rates, probability of transmission of a
pathogen, and duration of infectiousness. In rough terms, it measures
how many new infections each infections begets. The paper of Sy et al.
(2021) was selected for being, in the literature examined, almost alone
in supporting reproducible research. Accordingly, I wish to be clear
that my objective in singling their work for discussion is not to malign
their efforts, but rather to demonstrate how open and reproducible
research efforts can greatly help to accelerate discovery. More
concretely, open data and open code mean that an independent researcher
can, with only modest efforts, not only verify the findings reported,
but also examine the same data from a perspective which may not have
been available to the original researchers due to differences in
disciplinary perspectives, methodological traditions, and/or training,
among other possible factors. The example, which shows consequential
changes in the conclusions reached by different analyses, should serve
as a call to researchers to redouble their efforts to increase
transparency and reproducibility in their research. In this spirit, the
present paper also aims to show how data can be packaged in
well-documented, shareable units, and code can be embedded into
self-contained documents suitable for review and independent
verification. The source for this paper is an
\href{http://rmarkdown.rstudio.com}{R Markdown} document which, along
with the data package, are available in a public repository\footnote{\url{https://github.com/paezha/Reproductive-Rate-and-Density-US-Reanalyzed}}.

@@

The concern with population density and the spread of the virus during
the COVID-19 pandemic was fueled, at least in part, by dramatic scenes
seen in real-time around the world from large urban centers such as
Wuhan, Milan, Madrid, and New York. In theory, there are good reasons to
believe that higher density could have a positive association with the
transmission of a contagious virus. It has long been known that the
potential for inter-personal contact is greater in regions with higher
density (see for example the research on urban fields and
time-geography, including Farber and Páez, 2011; Moore, 1970; Moore and
Brown, 1970). Mathematically, models of exposure and contagion indicate
that higher densities can catalyze the transmission of contagious
diseases (Li et al., 2018; Rocklöv and Sjödin, 2020). The idea is
intuitive and likely at the root of messages, by some figures in
positions of authority, that regions with sparse population densities
faced lower risks from the pandemic\footnote{Governor Kristi Noem of
  South Dakota, for example, claimed that sparse population density
  allowed her state to face the pandemic down without the need for
  strict policy interventions
  \url{https://www.inforum.com/lifestyle/health/5025620-South-Dakota-is-not-New-York-City-Noem-defends-lack-of-statewide-COVID-19-restrictions}}.

As Rocklöv and Sjödin (Rocklöv and Sjödin, 2020) note, however,
mathematical models of contagion are valid at small-to-medium spaces
(and presumably, smaller time intervals too, such as time spent in
restaurants, concert halls, cruises), and the results do not necessarily
transfer to larger spatial units and longer time periods. There are
solid reasons for this: while in a restaurant, one can hardly avoid
being in proximity to other customers. On the other hand, a person can
choose to (or be forced to as a matter of policy) not go to a restaurant
in the first place. Nonetheless, the idea that high density correlates
with high transmission is so seemingly sensible that it is often taken
for granted even at the scale of large spaces (e.g., Cruz et al., 2020;
Micallef et al., 2020). In such conditions, however, there exists the
possibility of behavioral adaptations, which are difficult to capture in
the mechanistic framework of differential equations (or can be missing
in agent-based models, e.g., Gomez et al., 2021); these adaptations, in
fact, can be a key aspect of disease transmission.

A plausible behavioral adaptation during a pandemic, especially one
broadcast as widely and intensely as COVID-19, is risk compensation.
Risk compensation is a process whereby people adjust their behavior in
response to their \emph{perception} of risk (Noland, 1995; Phillips et
al., 2011; Richens et al., 2000). In the case of COVID-19, Chauhan et
al. (Chauhan et al., 2021) have found that perception of risks in the US
varies between rural, suburban, and urban residents, with rural
residents in general expressing less concern about the virus. It is
possible that people who listened to the message of leaders saying that
they were safe from the virus because of low density may not have taken
adequate precautions. Conversely, people in dense places who could more
directly observe the impact of the pandemic may have become overly
cautious. Both Paez et al. (2020) and Hamidi et al. (2020b) posit this
mechanism (i.e., greater compliance with social distancing in denser
regions) to explain the results of their analyses. The evidence
available does indeed show that there were important changes in behavior
with respect to mobility during the pandemic (Harris and Branion-Calles,
2021; Jamal and Paez, 2020; Molloy et al., 2020); furthermore, shelter
in place orders may have had greater buy-in from the public in higher
density regions (Feyman et al., 2020; Hamidi and Zandiatashbar, 2021),
and the associated behavior may have persisted beyond the duration of
official social-distancing policies (Praharaj et al., 2020). In
addition, there is evidence that changes in mobility correlated with the
trajectory of the pandemic (Noland, 2021; Paez, 2020). Given the
potential for behavioral adaptation, the question of density becomes
more nuanced: it is not just a matter of proximity, but also of human
behavior, which is better studied using population-level data and
models.

@@

On the one hand, there are studies that report positive associations
between population density and various COVID-19-related outcomes. Bhadra
(2021), for example, reported a moderate positive correlation between
the spread of COVID-19 and population density at the district level in
India, however their analysis was bivariate and did not control for
other variables, such as income. Similarly, Kadi and Khelfaoui (2020)
found a positive and significant correlation between number of cases and
population density in cities in Algeria in a series of simple regression
models (i.e., without other controls). A question in these relatively
simple analyses is whether density is not a proxy for other factors.
Other studies have included controls, such as Pequeno et al. (2020), a
team that reported a positive association between density and cumulative
counts of confirmed COVID-19 cases in state capitals in Brazil after
controlling for covariates, including income, transport connectivity,
and economic status. In a similar vein, Fielding-Miller et al. (2020)
reported a positive relationship between the absolute number of COVID-19
deaths and population density (rate) in rural counties in the US. Roy
and Ghosh (2020) used a battery of machine learning techniques to find
discriminatory factors, and a positive and significant association
between COVID-19 infection and death rates in US states. Wong and Li
(2020) also found a positive and significant association between
population density and number of confirmed COVID-19 cases in US
counties, using both univariate and multivariate regressions with
spatial effects. More recently, Sy et al. (2021) reported that the basic
reproductive number of COVID-19 in US counties tended to increase with
population density, but at a decreasing rate at higher densities.

On the flip side, a number of studies report non-significant or negative
associations between population density and COVID-19 outcomes. This
includes the research of Sun et al. (2020) who did not find evidence of
significant correlation between population density and confirmed number
of cases per day \emph{in conditions of lockdown} in China. This finding
echoes the results of Paez et al. (2020), who in their study of
provinces in Spain reported non-significant associations between
population density and infection rates in the early days of the first
wave of COVID-19, and negative significant associations in the later
part of the first lockdown. Similarly, Skórka et al. (2020) found zero
or negative associations between population density and infection
numbers/deaths by country. Fielding-Miller et al. (2020) contrast their
finding about rural counties with a negative relationship between
COVID-19 deaths and population density in urban counties in the US. For
their part, in their investigation of doubling time, White and
Hébert-Dufresne (2020) identified a negative and significant correlation
between population density and doubling time in US states. Likewise,
Khavarian-Garmsir et al. (2021) found a small negative (and significant)
association between population density and COVID-19 morbidity in
districts in Tehran. Finally, two of the most complete studies in the
US, by Hamidi et al. (2020a) and Hamidi et al. (2020b), used an
extensive set of controls to find negative and significant correlations
between density and COVID-19 cases and fatalities at the level of
counties in the US.

As can be seen, these studies are implemented at different scales in
different regions of the world. They also use a range of techniques,
from correlation analysis, to multivariate regression, spatial
regressions, and machine learning techniques. This is natural and to be
expected: individual researchers have only limited time and expertise.
This is why reproducibility is important. To pick an example (which will
be further elaborated in later sections of this paper), the study of Sy
et al. (2021), hereafter referred to as SWN, would immediately grab the
attention of a researcher with expertise in spatial analysis.

@@

There are various reasons why a project can fail to be reproducible. In
some cases, there might be legitimate reasons to withhold the data,
perhaps due to confidentiality and privacy reasons (e.g., Lee et al.,
2020). But in many other cases the data are publicly available, which in
fact has commonly been the case with population-level COVID-19
information. Typically the provenance of the data is documented, but in
numerous studies the data themselves are not shared (Amadu et al., 2021;
Bhadra et al., 2021; Cruz et al., 2020; Feng et al., 2020;
Fielding-Miller et al., 2020; Hamidi et al., 2020a, 2020b; Inbaraj et
al., 2021; Souris and Gonzalez, 2020). As any researcher can attest,
collecting, organizing, and preparing data for a project can take a
substantial amount of time. Pointing to the sources of data, even when
these sources are public, is a small step towards reproducibility-but
only a very small one. Faced with the prospect of having to recreate a
data set from raw sources is probably sufficient to dissuade all but the
most dedicated (or stubborn) researcher from independent verification.
This is true even if part of the data are shared (e.g., Wong and Li,
2020). In other cases, data are shared, but the processes followed in
the preparation of the data are not fully documented (Ahmad et al.,
2020; Skórka et al., 2020). These processes matter, as shown by the
errors in the spreadsheets of Reinhart and Rogoff (see Herndon et al.,
2014 for the discovery of these errors), as well as by the data of
biologist Jonathan Pruitt that led to an ``avalanche'' of paper
retractions (see Viglione, 2020). Another situation is when papers share
well-documented data, but fail to provide the code used in the analysis
(Noury et al., 2021; Pequeno et al., 2020; Wang et al., 2021). Making
code available only ``on demand'' (e.g., Brandtner et al., 2021) is an
unnecessary barrier when most journals offer the facility to share
supplemental materials online. Then there are those papers that more
closely comply with reproducibility standards, and share well-documented
processes and data, as well as the code used in any analyses reported
(Feyman et al., 2020; Paez et al., 2020; Stephens et al., 2021; Sy et
al., 2021; White and Hébert-Dufresne, 2020). Even in this case, the
pressure to publish ``new findings'' instead of replication studies can
act as a deterrent, perhaps particularly for younger
researchers\footnote{The present paper was desk rejected by three
  journals that had previously published research on population density
  and the spread of COVID-19; in one case, the paper was too opinionated
  for the journal, in the other two cases, the paper was not a ``good
  fit'' despite dealing with a nearly identical issue as papers
  previously published in said journals.}.

@@

A problematic issue with non-random sample selection is that parameter
estimates can become unreliable, and numerous techniques have been
developed to address this. A model useful for sample selection problems
is Heckman's selection model (see Maddala, 1983). The selection model is
in fact a system of two equations, as follows: \[
\begin{array}{c}
y_i^{S*} = \beta^{S\prime}x_i^S+\epsilon_i^S\\
y_i^{O*} = \beta^{O\prime}x_i^O+\epsilon_i^O
\end{array}
\] \noindent where \(y_i^{S*}\) is a latent variable for the sample
selection process, and \(y_i^{O*}\) is the latent outcome. Vectors
\(x_i^S\) and \(x_i^O\) are explanatory variables (with the possibility
that \(x_i^S = x_i^S\)). Both equations include random terms (i.e.,
\(\epsilon_i^S\) and \(\epsilon_i^O\)). The first equation is designed
to model the \emph{probability} of sampling, and the second equation the
outcome of interest (say \(R_0\)). The random terms are jointly
distributed and correlated with parameter \(\rho\).

@@

The second model in Table \ref{tab:selection-results} (Selection Model
2) changes the way the variables are entered into the model. The
log-transformation of density in SWN and Selection Model 1 assumes that
the association between density and \(R_0\) is monotonically increasing
(if the sign of the coefficient is positive) or decreasing (if the sign
of the coefficient is negative). There are some indications that the
relationship may actually not be monotonical. For example, Paez et al.
(2020) found a positive (if non-significant) relationship between
density and incidence of COVID-19 in the provinces of Spain at the
beginning of the pandemic. This changed to a negative (and significant)
relationship during the lockdown. In the case of the US, Fielding-Miller
et al. (2020) found that the association between COVID-19 deaths and
population density was positive in rural counties, but negative in urban
counties. A variable transformation that allows for non-monotonic
changes in the relationship is the square of the density.

>>>>

% tightlist command for lists without linebreak
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}

\begin{document}

\begin{frontmatter}

  \title{Reproducibility of research during COVID-19: examining the case
of population density and the basic reproductive rate from the
perspective of spatial analysis}
    \author[McMaster University]{Antonio Paez%
  \corref{cor1}%
  }
   \ead{paezha@mcmaster.ca} 
      \affiliation[McMaster University]{School of Earth, Environment and
Society, 1280 Main St West, Hamilton, Ontario L8S 4K1 Canada}
    \cortext[cor1]{Corresponding author}

@@

The emergence of the novel SARS-CoV-2 coronavirus in 2019, and the
global pandemic that followed in its wake, led to an explosive growth of
research around the globe. According to Fraser et al.
\citeyearpar{Fraser2021evolving}, over 125,000 COVID-19-related papers
were released in the first ten months from the first confirmed case of
the disease. Of these, more than 30,000 were shared in pre-print
servers, the use of which also exploded in the past year
\citep{Kwon2021swamped, Vlasschaert2020proliferation, Anazco2021publication}.

Given the ruinous human and economic cost of the pandemic, there has
been a natural tension in the scientific community between the need to
publish research results quickly and the imperative to maintain
consistently high quality standards in scientific reporting; indeed, a
call for maintaining the standards in published research termed the
deluge of COVID-19 publications a ``carnage of substandard research''
\citep{Bramstedt2020carnage}. Part of the challenge of maintaining
quality standards in published research is that, despite an abundance of
recommendations and guidelines
\citep[e.g.,][]{Broggini2017reproducible, Ince2012case, Ioannidis2014increasing, Brunsdon2020opening},
in practice reproducibility has remained a lofty and somewhat
aspirational goal
\citep{Konkol2019examination, Konkol2019computational}. As reported in
the literature, only a woefully small proportion of published research
was actually reproducible before the pandemic
\citep{Iqbal2016reproducible, Stodden2018empirical}, and the situation
does not appear to have changed substantially since
\citep{Sumner2020reproducibility, Gustot2020quality}.

The push for open software and data
\citep[e.g.,][]{Bivand2020progress, Arribas2021open}, along with more
strenuous efforts towards open, reproducible research, is simply a
continuation of long-standing scientific practices of independent
verification. Despite the (at times disproportionate) attention that
high profile scandals in science tend to elicit in the media, science as
a collective endeavor is remarkable for being a self-correcting
enterprise, one with built-in mechanisms and incentives to weed out
erroneous ideas. Over the long term, facts tend to prevail in science.
At stake is the shorter-term impacts that research may have in other
spheres of economic and social life. The case of economists Reinhart and
Rogoff comes to mind: by the time the inaccuracies and errors in their
research were uncovered \citep[see][]{Herndon2014high}, their claims
about debt and economic growth had already been seized by policy-makers
on both sides of the Atlantic to justify austerity policies in the
aftermath of the Great Recession of 2007-2009\footnote{Nobel Prize in
  Economics Paul Krugman noted that ``Reinhart--Rogoff may have had more
  immediate influence on public debate than any previous paper in the
  history of economics''
  \url{https://www.nybooks.com/articles/2013/06/06/how-case-austerity-has-crumbled/?pagination=false}}.
As later research has demonstrated, those policies cast a long shadow,
and their sequels continued to be felt for years \citep{Basu2017ten}.

In the context of COVID-19, a topic that has grabbed the imagination of
numerous thinkers has been the prospect of life in cities after the
pandemic \citep[e.g.,][]{Florida2020how}; as a result, the implications
of the pandemic for urban planning, design, and management are the topic
of ongoing research \citep[e.g.,][]{Sharifi2020covid}. The fact that the
worst of the pandemic was initially felt in dense population centers
such as Wuhan, Milan, Madrid, and New York, unleashed a torrent of
research into the associations between density and the spread of the
pandemic. The answers to some important questions hang on the results of
these research efforts. For example, are lower density regions safer
from the pandemic? Are de-densification policies warranted, even if just
in the short term? In the longer term, will the risks of life in high
density regions presage a flight from cities? And, what are the
implications of the pandemic for future urban planning and practice?
Over the past year, numerous papers have sought to throw light on the
underlying issue of density and the pandemic; nonetheless the results,
as will be detailed next, remain mixed. Further, to complicate matters,
precious few of these studies appear to be sufficiently open to support
independent verification.

The objective of this paper is to illustrate the importance of
reproducibility in research in the context of the flood of COVID-19
papers. For this, I focus on a recent study by Sy et al.
\citeyearpar{Sy2021population} that examined the correlation between the
basic reproductive number of COVID-19, \(R_0\), and population density.
The basic reproductive number is a summary measure of contact rates,
probability of transmission of a pathogen, and duration of
infectiousness. In rough terms, it measures how many new infections each
infections begets. The paper of Sy et al. \citeyearpar{Sy2021population}
was selected for being, in the literature examined, almost alone in
supporting reproducible research. Accordingly, I wish to be clear that
my objective in singling their work for discussion is not to malign
their efforts, but rather to demonstrate how open and reproducible
research efforts can greatly help to accelerate discovery. More
concretely, open data and open code mean that an independent researcher
can, with only modest efforts, not only verify the findings reported,
but also examine the same data from a perspective which may not have
been available to the original researchers due to differences in
disciplinary perspectives, methodological traditions, and/or training,
among other possible factors. The example, which shows consequential
changes in the conclusions reached by different analyses, should serve
as a call to researchers to redouble their efforts to increase
transparency and reproducibility in their research. In this spirit, the
present paper also aims to show how data can be packaged in
well-documented, shareable units, and code can be embedded into
self-contained documents suitable for review and independent
verification. The source for this paper is an
\href{http://rmarkdown.rstudio.com}{R Markdown} document which, along
with the data package, are available in a public repository\footnote{\url{https://github.com/paezha/Reproductive-Rate-and-Density-US-Reanalyzed}}.

@@

The concern with population density and the spread of the virus during
the COVID-19 pandemic was fueled, at least in part, by dramatic scenes
seen in real-time around the world from large urban centers such as
Wuhan, Milan, Madrid, and New York. In theory, there are good reasons to
believe that higher density could have a positive association with the
transmission of a contagious virus. It has long been known that the
potential for inter-personal contact is greater in regions with higher
density \citep[see for example the research on urban fields and
time-geography,
including][]{Farber2011running, Moore1970urban, Moore1970some}.
Mathematically, models of exposure and contagion indicate that higher
densities can catalyze the transmission of contagious diseases
\citep{Rocklov2020high, Li2018effect}. The idea is intuitive and likely
at the root of messages, by some figures in positions of authority, that
regions with sparse population densities faced lower risks from the
pandemic\footnote{Governor Kristi Noem of South Dakota, for example,
  claimed that sparse population density allowed her state to face the
  pandemic down without the need for strict policy interventions
  \url{https://www.inforum.com/lifestyle/health/5025620-South-Dakota-is-not-New-York-City-Noem-defends-lack-of-statewide-COVID-19-restrictions}}.

As Rocklöv and Sjödin \citep{Rocklov2020high} note, however,
mathematical models of contagion are valid at small-to-medium spaces
(and presumably, smaller time intervals too, such as time spent in
restaurants, concert halls, cruises), and the results do not necessarily
transfer to larger spatial units and longer time periods. There are
solid reasons for this: while in a restaurant, one can hardly avoid
being in proximity to other customers. On the other hand, a person can
choose to (or be forced to as a matter of policy) not go to a restaurant
in the first place. Nonetheless, the idea that high density correlates
with high transmission is so seemingly sensible that it is often taken
for granted even at the scale of large spaces
\citep[e.g.,][]{Cruz2020exploring, Micallef2020first}. In such
conditions, however, there exists the possibility of behavioral
adaptations, which are difficult to capture in the mechanistic framework
of differential equations \citep[or can be missing in agent-based
models, e.g.,][]{Gomez2021infekta}; these adaptations, in fact, can be a
key aspect of disease transmission.

A plausible behavioral adaptation during a pandemic, especially one
broadcast as widely and intensely as COVID-19, is risk compensation.
Risk compensation is a process whereby people adjust their behavior in
response to their \emph{perception} of risk
\citep{Noland1995perceived, Richens2000condoms, Phillips2011risk}. In
the case of COVID-19, Chauhan et al. \citep{Chauhan2021covid} have found
that perception of risks in the US varies between rural, suburban, and
urban residents, with rural residents in general expressing less concern
about the virus. It is possible that people who listened to the message
of leaders saying that they were safe from the virus because of low
density may not have taken adequate precautions. Conversely, people in
dense places who could more directly observe the impact of the pandemic
may have become overly cautious. Both Paez et al.
\citeyearpar{Paez2020spatio} and Hamidi et al.
\citeyearpar{Hamidi2020density} posit this mechanism (i.e., greater
compliance with social distancing in denser regions) to explain the
results of their analyses. The evidence available does indeed show that
there were important changes in behavior with respect to mobility during
the pandemic
\citep{Jamal2020Changes, Harris2021Changes, Molloy2020Tracing};
furthermore, shelter in place orders may have had greater buy-in from
the public in higher density regions
\citep{Feyman2020effectiveness, Hamidi2021compact}, and the associated
behavior may have persisted beyond the duration of official
social-distancing policies \citep{Praharaj2020Using}. In addition, there
is evidence that changes in mobility correlated with the trajectory of
the pandemic \citep{Paez2020using, Noland2021mobility}. Given the
potential for behavioral adaptation, the question of density becomes
more nuanced: it is not just a matter of proximity, but also of human
behavior, which is better studied using population-level data and
models.

@@

On the one hand, there are studies that report positive associations
between population density and various COVID-19-related outcomes. Bhadra
\citeyearpar{Bhadra2021impact}, for example, reported a moderate
positive correlation between the spread of COVID-19 and population
density at the district level in India, however their analysis was
bivariate and did not control for other variables, such as income.
Similarly, Kadi and Khelfaoui \citeyearpar{Kadi2020population} found a
positive and significant correlation between number of cases and
population density in cities in Algeria in a series of simple regression
models (i.e., without other controls). A question in these relatively
simple analyses is whether density is not a proxy for other factors.
Other studies have included controls, such as Pequeno et al.
\citeyearpar{Pequeno2020air}, a team that reported a positive
association between density and cumulative counts of confirmed COVID-19
cases in state capitals in Brazil after controlling for covariates,
including income, transport connectivity, and economic status. In a
similar vein, Fielding-Miller et al. \citeyearpar{Fielding2020social}
reported a positive relationship between the absolute number of COVID-19
deaths and population density (rate) in rural counties in the US. Roy
and Ghosh \citeyearpar{Roy2020factors} used a battery of machine
learning techniques to find discriminatory factors, and a positive and
significant association between COVID-19 infection and death rates in US
states. Wong and Li \citeyearpar{Wong2020spreading} also found a
positive and significant association between population density and
number of confirmed COVID-19 cases in US counties, using both univariate
and multivariate regressions with spatial effects. More recently, Sy et
al. \citeyearpar{Sy2021population} reported that the basic reproductive
number of COVID-19 in US counties tended to increase with population
density, but at a decreasing rate at higher densities.

On the flip side, a number of studies report non-significant or negative
associations between population density and COVID-19 outcomes. This
includes the research of Sun et al. \citeyearpar{Sun2020impacts} who did
not find evidence of significant correlation between population density
and confirmed number of cases per day \emph{in conditions of lockdown}
in China. This finding echoes the results of Paez et al.
\citeyearpar{Paez2020spatio}, who in their study of provinces in Spain
reported non-significant associations between population density and
infection rates in the early days of the first wave of COVID-19, and
negative significant associations in the later part of the first
lockdown. Similarly, \citet{Skorka2020macroecology} found zero or
negative associations between population density and infection
numbers/deaths by country. Fielding-Miller et al.
\citeyearpar{Fielding2020social} contrast their finding about rural
counties with a negative relationship between COVID-19 deaths and
population density in urban counties in the US. For their part, in their
investigation of doubling time, White and Hébert-Dufresne
\citeyearpar{White2020state} identified a negative and significant
correlation between population density and doubling time in US states.
Likewise, \citet{Khavarian2021high} found a small negative (and
significant) association between population density and COVID-19
morbidity in districts in Tehran. Finally, two of the most complete
studies in the US, by \citet{Hamidi2020longitudinal} and
\citet{Hamidi2020density}, used an extensive set of controls to find
negative and significant correlations between density and COVID-19 cases
and fatalities at the level of counties in the US.

As can be seen, these studies are implemented at different scales in
different regions of the world. They also use a range of techniques,
from correlation analysis, to multivariate regression, spatial
regressions, and machine learning techniques. This is natural and to be
expected: individual researchers have only limited time and expertise.
This is why reproducibility is important. To pick an example (which will
be further elaborated in later sections of this paper), the study of Sy
et al. \citeyearpar{Sy2021population}, hereafter referred to as SWN,
would immediately grab the attention of a researcher with expertise in
spatial analysis.

@@

There are various reasons why a project can fail to be reproducible. In
some cases, there might be legitimate reasons to withhold the data,
perhaps due to confidentiality and privacy reasons
\citep[e.g.,][]{Lee2020human}. But in many other cases the data are
publicly available, which in fact has commonly been the case with
population-level COVID-19 information. Typically the provenance of the
data is documented, but in numerous studies the data themselves are not
shared
\citep{Amadu2021assessing, Bhadra2021impact, Cruz2020exploring, Feng2020spread, Fielding2020social, Hamidi2020longitudinal, Hamidi2020density, Inbaraj2021seroprevalence, Souris2020covid}.
As any researcher can attest, collecting, organizing, and preparing data
for a project can take a substantial amount of time. Pointing to the
sources of data, even when these sources are public, is a small step
towards reproducibility-but only a very small one. Faced with the
prospect of having to recreate a data set from raw sources is probably
sufficient to dissuade all but the most dedicated (or stubborn)
researcher from independent verification. This is true even if part of
the data are shared \citep[e.g.,][]{Wong2020spreading}. In other cases,
data are shared, but the processes followed in the preparation of the
data are not fully documented
\citep{Ahmad2020association, Skorka2020macroecology}. These processes
matter, as shown by the errors in the spreadsheets of Reinhart and
Rogoff \citep[see][ for the discovery of these errors]{Herndon2014high},
as well as by the data of biologist Jonathan Pruitt that led to an
``avalanche'' of paper retractions \citep[see][]{Viglione2020avalanche}.
Another situation is when papers share well-documented data, but fail to
provide the code used in the analysis
\citep{Noury2021how, Pequeno2020air, Wang2021transmission}. Making code
available only ``on demand'' \citep[e.g.,][]{Brandtner2021creatures} is
an unnecessary barrier when most journals offer the facility to share
supplemental materials online. Then there are those papers that more
closely comply with reproducibility standards, and share well-documented
processes and data, as well as the code used in any analyses reported
\citep{Paez2020spatio, Feyman2020effectiveness, Stephens2021impact, White2020state, Sy2021population}.
Even in this case, the pressure to publish ``new findings'' instead of
replication studies can act as a deterrent, perhaps particularly for
younger researchers\footnote{The present paper was desk rejected by
  three journals that had previously published research on population
  density and the spread of COVID-19; in one case, the paper was too
  opinionated for the journal, in the other two cases, the paper was not
  a ``good fit'' despite dealing with a nearly identical issue as papers
  previously published in said journals.}.

@@

A problematic issue with non-random sample selection is that parameter
estimates can become unreliable, and numerous techniques have been
developed to address this. A model useful for sample selection problems
is Heckman's selection model \citep[see][]{Maddala1983limited}. The
selection model is in fact a system of two equations, as follows: \[
\begin{array}{c}
y_i^{S*} = \beta^{S\prime}x_i^S+\epsilon_i^S\\
y_i^{O*} = \beta^{O\prime}x_i^O+\epsilon_i^O
\end{array}
\] \noindent where \(y_i^{S*}\) is a latent variable for the sample
selection process, and \(y_i^{O*}\) is the latent outcome. Vectors
\(x_i^S\) and \(x_i^O\) are explanatory variables (with the possibility
that \(x_i^S = x_i^S\)). Both equations include random terms (i.e.,
\(\epsilon_i^S\) and \(\epsilon_i^O\)). The first equation is designed
to model the \emph{probability} of sampling, and the second equation the
outcome of interest (say \(R_0\)). The random terms are jointly
distributed and correlated with parameter \(\rho\).

@@

The second model in Table \ref{tab:selection-results} (Selection Model
2) changes the way the variables are entered into the model. The
log-transformation of density in SWN and Selection Model 1 assumes that
the association between density and \(R_0\) is monotonically increasing
(if the sign of the coefficient is positive) or decreasing (if the sign
of the coefficient is negative). There are some indications that the
relationship may actually not be monotonical. For example, Paez et al.
\citeyearpar{Paez2020spatio} found a positive (if non-significant)
relationship between density and incidence of COVID-19 in the provinces
of Spain at the beginning of the pandemic. This changed to a negative
(and significant) relationship during the lockdown. In the case of the
US, Fielding-Miller et al. \citeyearpar{Fielding2020social} found that
the association between COVID-19 deaths and population density was
positive in rural counties, but negative in urban counties. A variable
transformation that allows for non-monotonic changes in the relationship
is the square of the density.
